---
title: "Uber Data Analysis"
author: "Lucas Varela"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggthemes)
library(lubridate)
library(dplyr)
library(tidyr)
library(DT)
library(scales)
```

# Preparation

Data storytelling plays a crucial role in Machine Learning, allowing companies to grasp the context of different operations. Through visualization, companies can effectively comprehend intricate data, extracting insights that aid in decision-making. That's why I decided to do this data analysis project focused on Uber data from 2014.

### Load datasets

```{r echo=TRUE, warning=FALSE}
load("rda/apr_data.rda")
load("rda/may_data.rda")
load("rda/jun_data.rda")
load("rda/jul_data.rda")
load("rda/aug_data.rda")
load("rda/sep_data.rda")
```

### Create vector of colors for the plots

```{r echo=TRUE}
colors <- c("#CC1011", "#665555", "#05a399", "#cfcaca", "#f5e840", "#0683c9", "#e075b0")
```

### Formatting

Now I bind the different datasets and perform the appropriate formatting of date time columns, while also creating column factors of time objects like dat, month, year, etc.

```{r warning=FALSE}
data_2014 <- rbind(apr_data, may_data, jun_data, jul_data, aug_data, sep_data)

# Format date/time columns
data_2014$Date.Time <- as.POSIXct(data_2014$Date.Time, format = "%m/%d/%Y %H:%M:%S")
data_2014$Time <- format(as.POSIXct(data_2014$Date.Time, format = "%m/%d/%Y %H:%M:%S"), format = "%H:%M:%S")
data_2014$Date.Time <- ymd_hms(data_2014$Date.Time)

# Create columns factors for day, month, year, etc.
data_2014$day <- factor(day(data_2014$Date.Time))
data_2014$month <- factor(month(data_2014$Date.Time, label = TRUE))
data_2014$year <- factor(year(data_2014$Date.Time))
data_2014$dayofweek <- factor(wday(data_2014$Date.Time, label = TRUE))
data_2014$hour <- factor(hour(hms(data_2014$Time)))
data_2014$minute <- factor(minute(hms(data_2014$Time)))
data_2014$second <- factor(second(hms(data_2014$Time)))
```

# Visualization

In this step I'll use ggplot to visualize our wrangled data.

### Trips by the hours in a day

```{r echo=TRUE}
hour_data <- data_2014 %>%
	group_by(hour) %>%
	dplyr::summarize(Total = n())

ggplot(hour_data, aes(hour, Total)) + 
	geom_bar(stat = "identity", fill = "steelblue", color = "red") +
	ggtitle("Trips Every Hour") +
	theme(legend.position = "none") +
	scale_y_continuous(labels = comma)
```

### Trips by hour and month

```{r echo=TRUE}
month_hour <- data_2014 %>%
	group_by(month, hour) %>%
	dplyr::summarize(Total = n())

ggplot(month_hour, aes(hour, Total, fill = month)) + 
	geom_bar(stat = "identity") +
	ggtitle("Trips by Hour and Month") +
	scale_y_continuous(labels = comma)
```

### Trips every day

```{r echo=TRUE}
day_group <- data_2014 %>%
	group_by(day) %>%
	dplyr::summarize(Total = n())

ggplot(day_group, aes(day, Total)) + 
	geom_bar(stat = "identity", fill = "steelblue") +
	ggtitle("Trips Every Day") +
	theme(legend.position = "none") +
	scale_y_continuous(labels = comma)
```

### Trips by day and month

```{r echo=TRUE}
day_month_group <- data_2014 %>%
	group_by(month, day) %>%
	dplyr::summarize(Total = n())

ggplot(day_month_group, aes(day, Total, fill = month)) + 
	geom_bar(stat = "identity") +
	ggtitle("Trips by Day and Month") +
	scale_y_continuous(labels = comma) +
	scale_fill_manual(values = colors)
```

### Trips by month

```{r echo=TRUE}
month_group <- data_2014 %>%
	group_by(month) %>%
	dplyr::summarize(Total = n())

ggplot(month_group, aes(month, Total, fill = month)) + 
	geom_bar(stat = "identity") +
	ggtitle("Trips by Month") +
	theme(legend.position = "none") +
	scale_y_continuous(labels = comma) +
	scale_fill_manual(values = colors)
```

### Trips by month and weekday

```{r echo=TRUE}
month_weekday <- data_2014 %>%
	group_by(month, dayofweek) %>%
	dplyr::summarize(Total = n())

ggplot(month_weekday, aes(month, Total, fill = dayofweek)) + 
	geom_bar(stat = "identity", position = "dodge") +
	ggtitle("Trips by Day and Month") +
	scale_y_continuous(labels = comma) +
	scale_fill_manual(values = colors)
```

### Trips by base

```{r echo=TRUE}
ggplot(data_2014, aes(Base)) + 
	geom_bar(fill = "darkred") +
	scale_y_continuous(labels = comma) +
	ggtitle("Trips by Bases")
```

### Trips by base and month

```{r echo=TRUE}
ggplot(data_2014, aes(Base, fill = month)) + 
	geom_bar(position = "dodge") +
	scale_y_continuous(labels = comma) +
	ggtitle("Trips by Bases and Month") +
	scale_fill_manual(values = colors)
```

### Trips by base and day of week

```{r echo=TRUE}
ggplot(data_2014, aes(Base, fill = dayofweek)) + 
	geom_bar(position = "dodge") +
	scale_y_continuous(labels = comma) +
	ggtitle("Trips by Bases and DayofWeek") +
	scale_fill_manual(values = colors)
```

### Trips by day and hour

```{r echo=TRUE}
day_and_hour <- data_2014 %>%
	group_by(day, hour) %>%
	dplyr::summarize(Total = n())

ggplot(day_and_hour, aes(day, hour, fill = Total)) +
	geom_tile(color = "white") +
	ggtitle("Heat Map by Hour and Day")
```

### Heat maps for different combinations

```{r echo=TRUE}
ggplot(day_month_group, aes(day, month, fill = Total)) +
	geom_tile(color = "white") +
	ggtitle("Heat Map by Month and Day")

ggplot(month_weekday, aes(dayofweek, month, fill = Total)) +
	geom_tile(color = "white") +
	ggtitle("Heat Map by Month and Day of Week")
```

### Heat maps by month and bases, and bases and day of week

```{r echo=TRUE}
month_base <-  data_2014 %>%
	group_by(Base, month) %>%
	dplyr::summarize(Total = n()) 

day0fweek_bases <-  data_2014 %>%
	group_by(Base, dayofweek) %>%
	dplyr::summarize(Total = n()) 

ggplot(month_base, aes(Base, month, fill = Total)) +
	geom_tile(color = "white") +
	ggtitle("Heat Map by Month and Bases")

ggplot(day0fweek_bases, aes(Base, dayofweek, fill = Total)) +
	geom_tile(color = "white") +
	ggtitle("Heat Map by Bases and Day of Week")
```

### NYC map based on all Uber rides during 2014 (Apr-Sep)

```{r echo=TRUE}
min_lat <- 40.5774
max_lat <- 40.9176
min_long <- -74.15
max_long <- -73.7004

ggplot(data_2014, aes(x = Lon, y = Lat)) +
	geom_point(size = 1, color = "blue") +
	scale_x_continuous(limits = c(min_long, max_long)) +
	scale_y_continuous(limits = c(min_lat, max_lat)) +
	theme_map() +
	ggtitle("NYC MAP BASED ON UBER RIDES DURING 2014 (APR-SEP)")
```

### NYC map based on Uber rides during 2014 (Apr-Sep) by base

```{r echo=TRUE}
ggplot(data_2014, aes(x = Lon, y = Lat, color = Base)) +
	geom_point(size = 1) +
	scale_x_continuous(limits = c(min_long, max_long)) +
	scale_y_continuous(limits = c(min_lat, max_lat)) +
	theme_map() +
	ggtitle("NYC MAP BASED ON UBER RIDES DURING 2014 (APR-SEP) by BASE")
```


# Conclusion
In the conclusion of our Uber data analysis R project, we learned the art of crafting data visualizations. Utilizing tools such as ggplot2, we skillfully generated diverse visual representations spanning different time frames throughout the year. This enabled us to deduce the impact of time on customer trips. As a final step, we created a geo plot of New York, unraveling insights into how users embarked on trips from various starting points.